---
title: "visualize_results"
author: "Connor McCaffrey"
date: "2025-05-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**This file corresponds to the contents of Section 5.1: Exploratory Data Analysis in my final report.**

### Load Packages

```{r message=F, warning=F}
library(tidyverse)
library(GGally) 
```

### Load Data

```{r}
# load from file
clean <- read.csv("clean.csv")
```

### Plot 1: Pass Rate by Genre

Note that much of this code repeats what is done in the data cleaning file.

```{r}
# get vector of popular genres (plus "All")
major_genres <- bechdel %>%
  
    # add "All" to each movie's genre list
    mutate(genre = paste(genre, ", All", sep="")) %>%
  
    # expand rows for each genre value
    separate_rows(genre, sep = ", ") %>%
  
    # get all genres with count >= 10
    group_by(genre) %>%
    summarize(count = n()) %>%
    filter(count >= 10, genre != "NA") %>%
    pull(genre)

major_genres
```

```{r}
# calculate pass rate for each genre (plus "All")
df_genre <- bechdel %>%
  
    # add "All" to each movie's genre list
    mutate(genre = paste(genre, ", All", sep=""),
           outcome = ifelse(binary == "PASS", 1, 0)) %>%
  
    # expand rows for each genre value
    separate_rows(genre, sep = ", ") %>%
  
    # set uncommon genres to "Other"
    mutate(genre = ifelse(genre %in% major_genres, genre, "Other")) %>%
  
    # calculate pass rate for each genre
    group_by(genre) %>%
    summarize(genre_rate = mean(outcome),
              count = n()) %>%
    
    # set colors for bar chart
    mutate(color = ifelse(genre == "All", "gray", "white"),
           genre = as.factor(genre)) %>%
    arrange(genre_rate)

df_genre
```

```{r}
# create bar chart
plot1 <- ggplot(df_genre, aes(y=reorder(genre, -genre_rate), x=genre_rate, fill=color)) +
    
    # add bars
    geom_bar(stat="identity", color="black", show.legend=F) +
  
    # add count labels
    geom_text(aes(label=count), x=0.07, hjust=1, size=3.5) + 
  
    # set colors
    scale_fill_manual(values=c("firebrick1", "gray90")) +
  
    # set x-axis breaks
    scale_x_continuous(breaks=seq(0, 0.8, 0.1)) +
  
    # set plot labels
    labs(title="Bechdel Test Pass Rate & Count by Genre",
         y="Genre",
         x="Pass Rate") +
  
    # set theme
    theme_bw() + theme(panel.grid.major.y = element_blank(),
                       plot.title = element_text(hjust=0.5))

plot1
```

### Plot 2: Pass Rate by Year

```{r}
# calculate pass rate by year
df_year <- clean %>%
    mutate(outcome = ifelse(binary == "PASS", 1, 0)) %>%
    group_by(year) %>%
    summarize(rate = mean(outcome),
              count = n())

df_year
```

```{r}
# create line chart
plot2 <- ggplot(df_year, aes(x=year, y=rate)) +
  
    # add lines
    geom_line(color="darkgray", linetype=1) +
  
    # add points
    geom_point(shape=21, fill="white", color="darkgray") +
  
    # set plot labels
    labs(title="Bechdel Test Pass Rate Over Time",
         y="Pass Rate",
         x="Year") +
  
    # set theme
    theme_bw() + theme(panel.grid.major.x = element_blank(),
                       panel.grid.minor.x = element_blank(),
                       plot.title = element_text(hjust=0.5))

plot2
```

### Plot 3: Scatterplot Matrix

```{r}
# theme settings
RED <- "maroon3"
BLUE <- "skyblue2"
DARKGRAY <- "gray30"
LIGHTGRAY <- "gray95"

# variable labels
var_labels = c("year" = "Year",
               "budget" = "Budget",
               "domgross" = "Dom. Gross",
               "intgross" = "Int. Gross",
               "metascore" = "Metascore",
               "imdb_rating" = "IMDb Rating",
               "runtime" = "Runtime",
               "imdb_votes" = "IMDb Votes")
```

```{r}
# function to draw balanced sample from outcome indicator
samp <- function(n, label) {
  pass <- sample(which(label == 1), n/2)
  fail <- sample(which(label == 0), n/2)
  c(pass, fail)
}
```

```{r}
# draw balanced sample from continuous variables only
df_scatter <- clean %>%
    mutate(outcome = factor(outcome, levels=c("0", "1"), labels=c("0", "1"))) %>%
  
    # select only continuous variables
    select(outcome, year, budget, domgross, intgross,
           metascore, imdb_rating, runtime, imdb_votes) %>%
  
    # drop rows with NA values
    na.omit() %>%
  
    # draw balanced sample of size 1000
    slice(sample(samp(1000, outcome)))
```

```{r}
# function to generate scatterplot in ggpairs panel
upperfun <- function(data, mapping) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.3, size=1) +
    scale_color_manual(values = c("0" = RED, "1" = BLUE))
}
```

Calling ggpairs() will spam this warning: "Adding another scale for colour, which will replace the existing scale.
Scale for colour is already present." Just ignore it.

```{r}
# create scatterplot matrix
plot3 <- ggpairs(df_scatter, aes(color=outcome, group=outcome, fill=outcome, alpha=0.05),
                 columns = 2:ncol(df_scatter),
                 lower = "blank",
                 upper = list(continuous = wrap(upperfun)),
                 labeller = as_labeller(var_labels)) +
  
  # set color and fill values
  scale_color_manual(values = c("0" = RED, "1" = BLUE)) +
  scale_fill_manual(values = c("0" = RED, "1" = BLUE)) +
  
  # set theme
  theme_bw() + theme(plot.title = element_text(lineheight = 1.1, hjust=0.5),
                     axis.text = element_blank(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_line(color = LIGHTGRAY),
                     panel.grid.minor = element_line(color = LIGHTGRAY))

plot3
```